---
title: "Final Project"
author: "Alan Qiao, Jacqueline Dinh"
format: pdf
---

## Read in the data

```{r load-libraries, message = F, echo = F}
library(readxl)
library(tidyverse)
library(broom)
library(Stat2Data)
library(tidymodels)
library(dplyr)
library(data.table)
library(tidyr)
```

```{r read-data, echo = F}

nc_school <- read_excel("NC_School_Performance.xlsx")
```

## Introduction

Through our project, we hope to predict grade level proficiency of a school across all EOG subjects based on demographics of the school using predictors such as total number of female, male, American Indian, Asian, Black, Hispanic, two or more races, white, economically disadvantaged, limited English proficient, and academically or intellectually gifted students, along with the total number of students and the number of students with disabilities. We intend to create a linear regression model and will calculate the predictors by manipulating the dataset.

Our EDA will include a plot of percentages of each demographic that is proficient to illustrate disparities and box plots categorized by school size.

Our hypothesis is that schools with higher percentages of white students will, on average, have a greater percentage of students that are grade level proficient. In addition, we hypothesize that a larger school will, on average, have a greater variability in students that are grade level proficient.

## Data 

To investigate our research question, we used the dataset NC School Performance (2016 - 2017). This dataset is taken from Durham Open Data, a program managed jointly by the City of Durham's Technology Solutions Department and the County of Durham's Information Services & Technology Department to make data open and free to everyone. Durham Open Data collected this data from the North Carolina Department of Public Instruction and manipulated the annual testing data of public schools in North Carolina from the 2016-2017 school year. The main purpose of this dataset is to record the percentage of students of various races and other factors described that are grade level proficient and college/career ready. In this dataset, there are 32 variables and 85,766 observations separated by school, subject, and standard. The variable Standard was separated into Grade Level Proficient and College and Career Ready, determined by students’ test scores of Level 4 & 5 on the CCR for college/career readiness and Level 3 & Above for grade level proficiency. The variable subject included all EOG/EOC Subjects and individual EOC Subjects along with the Math, Science, and Reading EOG and EOC assessments for each grade. To narrow the dataset down and compare schools, we chose to filter the dataset to view scores of the observations that were determined by the subject of all EOG/EOC Subjects and the standard of Grade Level Proficient. 

The dataset presented percentages that are greater than 95 as “>95” and percentages less than 5 as “<5.” To make sure our model does not skip over these observations, we manipulated the dataset to change these percentages to 97.5 and 2.5, respectively, as they are the averages of 95-100 and 0-5. Also, in the dataset, if the number of students that met a condition were less than 10, it would just display “<10.” We then manipulated the dataset again to change these counts to 5, the average of 0 and 10. In the dataset, a * indicates a school does not have tested grades/sufficient data for reporting, and these were changed to NA’s in our model. Rows that did not have sufficient data for the 'All Students' subgroup are excluded from this file. In addition, we didn’t include the demographic of American Indian in our model, as there was too little data present in the file because of how small their population is. Finally, we filtered out the State of North Carolina, which was listed as a school, as this was the average of all the schools present and we wanted to investigate individual school demographics.

In our manipulated dataset, to look at and compare an overall assessment for all public NC schools, we included the percentage and number of students that are grade level proficient in all EOG/EOC subjects for all, female, male, American Indian, Asian, Black, two or more races, white, economically disadvantaged (EDS), limited English proficient, disabled, and academically or intellectually gifted students for each school.


```{r clean-data, warning = F, echo = F}

nc_school <- setnames(nc_school, old = c('District Name', 'School Code', 'School Name', 
'SBE District', 'Subject', 'Standard (CCR - Level 4 & 5, GLP - Level 3 & Above)',
'Percentage of: All Students', 'Percentage of: Female', 'Percentage of:  Male', 
'Percentage of: American Indian', 'Percentage of: Asian', 'Percentage of: Black',
'Percentage of: Hispanic', 'Percentage of: Two or More Races', 
'Percentage of:  White', 'Percentage of: EDS', 'Percentage of: LEP', 
'Percentage of: SWD', 'Percentage of: AIG', 'Number of: All Students', 
'Number of: Female', 'Number of:  Male', 'Number of: American Indian', 
'Number of: Asian', 'Number of: Black','Number of: Hispanic', 
'Number of: Two or More Races', 'Number of:  White', 'Number of: EDS', 
'Number of: LEP', 'Number of: SWD', 'Number of: AIG'), new = c("district", 
"school_code", "school_name", "SBE", "subject", "standard", "all", "female", 
"male", "indigenous", "asian", "black_", "hispanic", "multi", "white_", "EDS", 
"LEP", "SWD", "AIG", "total_all", "total_female", "total_male", 
"total_indigenous", "total_asian", "total_black", "total_hispanic", 
"total_multi", "total_white", "total_EDS", "total_LEP", "total_SWD", 
"total_AIG"), skip_absent=TRUE)


nc_school <- nc_school |>
  mutate(across(everything(), ~ if_else(. == ">95", "97.5", .)))

nc_school <- nc_school |>
  mutate(across(everything(), ~ if_else(. == "<5", "2.5", .)))

nc_school <- nc_school |>
  mutate(across(everything(), ~ if_else(. == "<10", "5", .)))

nc_school <- nc_school |> 
  mutate(across(.cols = -c(1, 2, 3, 4, 5, 6), .fns = as.numeric))

nc_school1 <- nc_school |>
  filter(district == "State of North Carolina", subject == "All EOG/EOC Subjects",
         standard == "Grade Level Proficient")

nc_school1 <- nc_school1 |>
  mutate(across(everything(), as.character))

nc_school1<- nc_school1 |>
  pivot_longer(everything(), names_to = "variable", values_to = "value")


nc_school <- nc_school |>
  filter(!district == "State of North Carolina", subject == "All EOG/EOC Subjects",
         standard == "Grade Level Proficient")

nc_school <- nc_school |>
  filter(!all == 0)
```

```{r calculate-predictors, echo = F}

nc_school <- nc_school |>
  mutate(
    all_pop = total_all/(all*.01),
    female_pop = total_female/(female*.01),
    male_pop = total_male/(male*.01),
    asian_pop = total_asian/(asian*.01),
    black_pop = total_black/(black_*.01),
    hispanic_pop = total_hispanic/(hispanic*.01),
    multi_pop = total_multi/(multi*.01),
    white_pop = total_white/(white_*.01),
    EDS_pop = total_EDS/(EDS*.01),
    LEP_pop = total_LEP/(LEP*.01),
    SWD_pop = total_SWD/(SWD*.01),
    AIG_pop = total_AIG/(AIG*.01)
    )

nc_school <- nc_school |>
  mutate(
    female_dem = female_pop/all_pop * 100,
    male_dem = male_pop/all_pop * 100,
    asian_dem = asian_pop/all_pop * 100,
    black_dem = black_pop/all_pop * 100,
    hispanic_dem = hispanic_pop/all_pop * 100,
    multi_dem = multi_pop/all_pop * 100,
    white_dem = white_pop/all_pop * 100,
    EDS_dem = EDS_pop/all_pop * 100,
    LEP_dem = LEP_pop/all_pop * 100,
    SWD_dem = SWD_pop/all_pop * 100,
    AIG_dem = AIG_pop/all_pop * 100
    )

quantiles <- quantile(nc_school$white_dem, probs = c(1/3, 2/3), na.rm = TRUE)

quantiles

nc_school <- nc_school |>
  mutate(
    white_comp = case_when(
      white_dem < 29.98547 ~ "minority white",
      white_dem > 29.98547 & white_dem < 61.69422  ~ "bmid white",
      white_dem > 61.69422 ~ "amajority white"
      )
  )
```



```{r model, echo = F}

schools <- lm(all ~ female_dem + male_dem + asian_dem + black_dem + hispanic_dem + white_dem + EDS_dem + LEP_dem + SWD_dem + AIG_dem + white_comp, data = nc_school)

summary(schools)
```


```{r barplot}

nc_school1 <- nc_school1 |>
  filter(variable %in% c("all", "female", "male", "asian", "black_", "hispanic", "white_", "EDS", "LEP", "SWD", "AIG"))

ggplot(nc_school1, aes(x = variable, y = value)) +
  geom_bar(stat = "identity")
```

```{r residuals}

schools_aug <- augment(schools)
schools_aug |>
  ggplot(
    aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "darkred") +
  labs(x = "Fitted (predicted) value",
       y = "Residuals",
       title = "Fairly Random Scatter in Residual Plot")
```


```{r QQ-plot}
schools_aug |>
  ggplot(
    aes(sample= .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical Quantiles",
       y = "Sample quantiles",
       title = "Low Deviation for QQ Plot")
```

